---
title: "R Setup"
author: "Steven Delean (based on notes by Steve Pederson)"
date: "24th July 2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Installation Instructions {.tabset}

There are two primary pieces of software we'll need to install in preparation for the course: `R` and `RStudio`.
However, there are some differences in installation across Mac (OSX) and Windows, so please check the appropriate section below.
You will need to be connected to the internet during installation.

### OSX

1. **The [R Software](https://cran.r-project.org/bin/macosx/) itself.**
    + The latest release is 4.3.1, however any version >4.0 will be suitable. 
    + If you have High Sierra or lower (OSX $\leq$ 10.11) you will be restricted to R 4.2.3. If you have El Capitan or lower (OSX $\leq$ 10.11) you will be restricted to R 3.6.3. Whilst these will be mostly compatabile with today's material, you may like to consider upgrading your OS relatively soon. Bioinformatics tools are often developed quickly and may render this OS as more and more of a problem as time moves on.
2. **The IDE [R Studio](https://www.rstudio.com/products/rstudio/download/#download)** which is the software we use to interact with R
3. R is also dependent on [XQuartz](https://www.xquartz.org/) so please ensure you have the latest version

### Windows 

If you are running your own laptop where *you are the administrator*, installation is relatively easy.
Please install the following software in order:

1. **The [R Software](https://cran.r-project.org/bin/windows/base/) itself.** The latest release is 4.3.1, however any version >3.6 will be suitable
2. The IDE [R Studio](https://www.rstudio.com/products/rstudio/download/#download) which is the software we use to interact with R

### Ubuntu

Open a terminal (`Ctrl + Alt +T`).
Package for the current R 4.2. release* are available for most stable Desktop releases of Ubuntu until their official end of life date. However, only the latest Long Term Support (LTS) release is fully supported. As of May 2, 2022 the supported releases are

    Jammy Jellyfish (22.04, amd64 only)
    Impish Indri (21.10, amd64 only),
    Focal Fossa (20.04; LTS and amd64 only),
    Bionic Beaver (18.04; LTS), and
    Xenial Xerus (16.04; LTS).

Run these lines (if root, remove sudo) to tell Ubuntu about the R binaries at CRAN.

```
# update indices
sudo apt update -qq
# install two helper packages we need
sudo apt install --no-install-recommends software-properties-common dirmngr
# add the signing key (by Michael Rutter) for these repos
# To verify key, run gpg --show-keys /etc/apt/trusted.gpg.d/cran_ubuntu_key.asc 
# Fingerprint: E298A3A825C0D65DFD57CBB651716619E084DAB9
wget -qO- https://cloud.r-project.org/bin/linux/ubuntu/marutter_pubkey.asc | sudo tee -a /etc/apt/trusted.gpg.d/cran_ubuntu_key.asc
# add the R 4.0 repo from CRAN -- adjust 'focal' to 'groovy' or 'bionic' as needed
sudo add-apt-repository "deb https://cloud.r-project.org/bin/linux/ubuntu $(lsb_release -cs)-cran40/"
```

Here we use lsb_release -cs to access which Ubuntu flavor you run: one of “jammy”, impish”, “focal”, “bionic”, …

Then run
```
sudo apt install --no-install-recommends r-base
```
to install R and its dependencies.

For full instructions, see https://cran.r-project.org/bin/linux/ubuntu/

##

Once you have this all installed, open `RStudio` and look for the frame which is titled **`Console`**. 
Paste the following lines all at once, followed by **Enter**. If you are prompted to update any packages, type  `a` for all.

```{r, eval=FALSE}
pkgs <- c("tidyverse", "pander", "plotly", "ggrepel", "gridExtra", 
          "knitr", "rmarkdown", "readxl", "devtools", "kableExtra")
install.packages(pkgs, dependencies = TRUE)
```

This will install the required packages for the course.
These may take a while to install.
If you see an error which indicates a failure at the end, please ask a demonstrator to help

[Back to main page](./)